---
title: "Species Distribution Modelling with MaxEnt in R"
output: html_notebook
author: "Dony Indiarto"
---

***

##Chapter 1 Introduction

Synonyms: Climate modelling envelope, habitat modelling, and niche-modeling.

Main Aim: to estimate the similarity of the condition at any site to the conditions at any sites to the conditions at the locations of known occurence of a phenomenon.

##### Major steps in SDMs:

1. a compilation of species occurence locations
2. values of environmental predictor variables (such as climate), extracted from climate databases
3. model fitting
4. prediction of variable of interest across an the region of interest

Required packages:
__install.packages(c(’raster’, ’rgdal’, ’dismo’, ’rJava’))__

***
##Chapter 2 Species occurence data


All model perform better with  unbiased and error-free data. Garbage in garbage out applies.

> Improving the quality of occurrence data first, model comparison second.


```{r include=T, echo=T, eval=FALSE}
#load libraries
library(raster)
library(rgdal)
library(dismo)
library(rJava)
library(readr)
```

### 2.1 Presence data input arrangements
```{r Bradypus data example collapse=T}
#load occurence dataset
 bradypus <- read_csv(paste(system.file(package="dismo"), "/ex/bradypus.csv", sep=""))
head(bradypus)

```

Species occurence dataframe consists of:

1. species name
2. x or longitude
3. y or latitude

Do not change the order. A single row represents an occurence. Many occurence has 
Data can be gathered from Global Biodiversity Inventory Facility (GBIF) (http:
//www.gbif.org/).
~~We can use a function from dismo package to download species occurence data from GBIF :~~


Dismo uses the old GBIF API therefore use occ function from spocc package (Interface to many species occurrence data sources):

### 2.2 Load data from GBIF
```{r load data from GBIF}
library(spocc)
acaule<- occ(query="Solanum acaule", from='gbif', limit=10000)
acaule<-acaule$gbif$data$Solanum_acaule
#data dimension
dim(acaule)

#data column names
colnames(acaule)
library(plyr)
acaule$country<-revalue(acaule$country, c("Bolivia, Plurinational State of"="Bolivia"))

```


### 2.3 Data cleaning
#### 2.3.1 Unfiltered data visualisation
Create a map of species occurence

```{r unfiltered presence data}
#data filtering, select the records that have longitude and latitude data
acgeo <- (subset(acaule, !is.na(longitude) & !is.na(latitude)))
dim(acgeo)

library(maptools)
data("wrld_simpl")
plot(wrld_simpl, xlim=c(-40,20), ylim=c(-70,70), axes=T, col='light yellow')
box()
points(acgeo$longitude, acgeo$latitude, col='orange', pch=20, cex=0.75)
 points(acgeo$longitude, acgeo$latitude, col='red', cex=0.75)
 




```

*Solanum acaule* presents in the higher parts of the Andes mountains of southern Peru, Bolivia, and Northern Argentina. 

As we see from the map, there are some errors occured. Points in Sweden, Mexico, Southern argentina, and Brazil might not be true. Background knowledge on species distribution is required.

Common mistakes:

1. missing minus signs
2. duplicates (identical data points that were submitted twice or more)
3. latitude/longitude typo
4. 0 coordinates where 'NAs' were intended (not georeferenced yet)

#### 2.3.2 Filtered data visualisation based on expert judgement/background knowledge
```{r filtered presence data}

dim(acgeo) #unfiltered dataset
#remove error data pouints from acgeo
filtered_acgeo<-dplyr::filter(acgeo, (longitude<(-55)&latitude<0& latitude>=-35)) #removing data points on South Argentina, Peru, Sweden, and Brazil

dim(filtered_acgeo) #filtered dataset

#filtered dataset
plot(wrld_simpl, xlim=c(-40,20), ylim=c(-70,70), axes=T, col='light yellow')
box()
points(filtered_acgeo$longitude, filtered_acgeo$latitude, col='orange', pch=20, cex=0.75)
 points(filtered_acgeo$longitude, filtered_acgeo$latitude, col='red', cex=0.75)

```
#### 2.3.4 Cross-checking

cross-check corrdinates by visual and other means. In this case, we compare the country as specified by the records with the country implied by the coordinates.

```{r}
acg<-filtered_acgeo
library(sp)
coordinates(acg)<- ~longitude+latitude

gadm_countries<-readRDS("./data/countries_gadm28.rds")

crs(acg)<-crs(gadm_countries)

# a check whether both datasets are under same data type and crs
class (acg)
class(gadm_countries)

ovr <- over(acg, gadm_countries)
cntr <- ovr$NAME
i <- which(is.na(cntr)) # which data points do not match any country

j <- which(cntr!= acg$country) # which data points missmatch
cbind(as.character(cntr), acg$country)[j,]

plot(acg)
plot(gadm_countries, add=T, border='blue', lwd=2)
points(acg[j, ], col='red', pch=20, cex=2)


```

No data missmatch


##Chapter 3 Environmental variables